Certified Knowledge Compilation \\ with Application to Verified Model Counting \\ Experimental Details

### 2023-08-28 doc/JAIR/jair-data.tex
Completed data for structured vs. monolithic

Paper describes two ways to generate the forward implication proof:
structured and monolithic.  Can also combine the two: at any point in
the traversal during structured generation, the forward implication
proof for the entire subgraph can be generated by a single call to a proof-generating
SAT solver.  We'll refer to this as a \emph{hybrid} approach.

Ran some experiments on all but the largest benchmarks (presented
later) seeing under what conditions was it best to use a monolithic,
structured, or hybrid approach, and where the hybrid approach should
shift from structured to monolithic.

General findings:
\begin{itemize}
\item The tree ratio provides a reasonable discriminator for monolithic vs.~structured.  Based on experiments, set threshold at 5.0. Use monolithic below this and structured above.
\item We could not find any way to exploit the hybrid approach for the smaller benchmarks.
\item On the other hand, the monolithic approach becomes impractical beyond around $10^6$ defining clauses.  Even the best SAT solvers become unstable beyond this.
\item Therefore, use hybrid approach for very large problems with tree ratios below 5.0, shift to monolithic once the tree size is at most $10^6$.
\item Also tried with limit of $10^5$ clauses.  Generally did worse in terms of completions.
\end{itemize}

Our summary data is based on this set of rules.

\begin{itemize}
\item Of the 123 benchmarks, 110 were completed using the hybrid approach within 10,000 seconds.
\item One additional one completed within 10,000 for structural (and for hybrid with a threshold of $10^5$ clauses.
\item For seven others, was able to do one-sided proof
\item Only one had more than $10^9$ defining clauses.
\item Total of 5 (including 1 overflow case) for which no form of validation succeeded.
  \begin{itemize}
  \item Three of these had more than $10^9$ defining clauses
  \item Smallest had 405 million

Figure~\ref{fig:d4:cpog} compares the running time for the combination
of generating and checking the CPOG file (Y axis), versus the running
time for C4 (X Axis).  If we compare the ratio between the two we find the following:
\begin{itemize}
\item Min = 0.27 (i.e., Verification chain was $3.64\times$ faster
\item Max = 177.0
\item Median = 12.5
\end{itemize}

@@

Our first set of experiments applies full monolithic and full structured generation to the reduced benchmark set.  
After some experimenting, we devised the following ``hybrid'' method: when the tree ratio is at most $5.0$, use the monolithic approach, otherwise use the structured approach.
We set a time limit of 1000 seconds for the proof generation and checking.
Overall, our results on the 92 benchmarks were:
\begin{itemize}
\item Monolithic: 83 completed
\item Structured: 82 completed
\item Hybrid: 85 completed
\end{itemize}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 0.70]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
      legend style={at={(0.30,0.96)}},
      legend cell align={left},
                              x post scale=1.8, y post scale=2.0,
                              xmode=log,xmin=0.01,xmax=1000,
                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
                              ymin=0.01, ymax=2000,
                              ytick={0.01, 0.1,1.0,10,100,1000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
                              xlabel={D4 runtime (seconds)}, ylabel={Toolchain runtime (seconds)},
            ]

    \input{data-formatted/sub10M-seconds-hybrid}
    \legend{
      \scriptsize \textsf{Hybrid},
    }

    \input{data-formatted/sub10M-seconds-failures-hybrid}

          \end{axis}
\end{tikzpicture}
} % centering
\caption{Total runtime for toolchain using three different approaches to forward-implication proof generation, as function of D4 runtime}
\label{fig:d4:cpog:compare}
\end{figure}

@@

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 0.90]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log, ymin=100, ymax=1000000000, 
                              x post scale=1.8, y post scale=1.8,
                              xmin=0,xmax=1000,
                              xtick = {0,100,200,300,400,500,600,700,800,900,1000},
                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              xlabel={D4 runtime (seconds)}, ylabel={Defining Clauses},
%                              title={D4 Defining Clause Generation}
            ]
    \input{data-formatted/time+defining-success}
    \input{data-formatted/time+defining-failure}
          \end{axis}
\end{tikzpicture}
} % centering
\caption{Number of defining clauses in generated POG as function of D4 runtime (X linear scale)}
\label{fig:generate:d4}
\end{figure}

>>>>

@@

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 0.70]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
      legend style={at={(0.30,0.96)}},
      legend cell align={left},
                              x post scale=2.0, y post scale=2.0,
                              xmode=log,xmin=0.01,xmax=1800,
                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
                              ymode=log, ymin=0.01, ymax=1800,
                              ytick={0.01, 0.1,1.0,10,100,1000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
                              xlabel={Monolithic generation + check (seconds)}, ylabel={Structural generation + check (seconds)},
            ]

    \input{data-formatted/sub10M-seconds-subthresh}
    \input{data-formatted/sub10M-seconds-supthresh}
    \legend{
      \scriptsize \textsf{Tree ratio $\leq \; 5{.}0$},
      \scriptsize \textsf{Tree ratio $> \; 5{.}0$},
    }
    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (1000,1000)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,1000) (1000,1000)};
    \addplot[mark=none, color=lightblue] coordinates{(1000,0.01) (1000,1000)};
    \addplot[color=black,dashed] coordinates{(1e-2,1500) (1500,1500)};
    \addplot[color=black,dashed] coordinates{(1500,1e-2) (1500,1500)};

          \end{axis}
\end{tikzpicture}
} % centering
\caption{Structural (Y-axis) versus monolithic (X-axis) execution times.  Monolithic generally performed better for lower tree ratios.}
\label{fig:seconds:mono:structured}
\end{figure}

Our first set of experiments applies full monolithic and full
structured generation to the reduced benchmark set.  Figure
~\ref{fig:seconds:mono:structured} shows a plot comparing the two
approaches.  Each axis shows the number of seconds to generate and
check a CPOG file.  Data points to the left of the diagonal line ran
faster with the structural method, while those to the right ran faster
with the monolithic method.  Five of the benchmarks exceeded the 1000-second threshold for both approaches.
The data are divided into two groups:
those having tree ratios below 5.0 and those having tree ratios above
5.0.  
As can be seen there is some correlation.  Of the 85 benchmarks for which at least one completed within 1000 seconds:
\begin{itemize}
\item For those with tree ratios below 5.0, 12 ran faster with monolithic generation and 23 with structural
\item For those with tree ratios above 5.0, 37 ran faster with monolithic generation and 13 with structural
\end{itemize}
Based on this, we devised the following selection rule: when the
tree ratio is at most $5.0$, use the monolithic approach, otherwise
use the structured approach.  That would yield the better choice for
60 of the 85 cases.  In terms of failures, there is one benchmark with
a high tree ratio for which the monolithic approach requires 835
seconds, and the structural approach requires 1388. This is therefore
marked as a failure by following our rule.  On the other hand, the rule makes the better choice for four of the cases where one approach times out while the other completes.
Our data set was too sparse to do more tuning, including the optimal threshold selection.

Overall, our results on the 90 benchmarks were:
\begin{itemize}
\item Monolithic: 83 completed
\item Structured: 82 completed
\item Threshold of 5.0: 85 completed
\end{itemize}

We tried various experiments where the proof generator starts at the
top using a structural approach and then switches to a monolithic
approach once the tree size for a node drops below some threshold.  We
could not find any selection procedure that improved the overall
performance for the 90 smaller benchmarks.  On the other hand, it
proved useful for the very large benchmarks, because calling a SAT
solver on a problem with more than $10^6$ clauses often causes the
solver to fail.  We therefore refined the rule into a \emph{hybrid} approach to operate as follows:
\begin{enumerate}
\item Label all nodes by the tree size of the subgraph for which it is the root.
\item Compute the total size of the graph and the tree ratio of the root.
\item Proceed with proof generation with the following rules
  \begin{enumerate}
  \item If the tree ratio is below 5.0 and the tree size of the root is below $10^6$, do the entire proof generation with a monolithic approach
  \item If the tree ratio is below 5.0 and the tree size of the root is above $10^6$, start with a structural approach and shift to a monolithic approach once the tree size at a node is below $10^6$.
  \item If the tree ratio is above 5.0, do the entire proof generation
    with a structural approach.
  \end{enumerate}
\end{enumerate}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 0.70]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
      legend style={at={(0.30,0.96)}},
      legend cell align={left},
                              x post scale=2.0, y post scale=2.0,
                              xmode=log,xmin=1e2,xmax=3e7,
                              xtick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000, 1e10}, 
                              xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$, $10^{10}$},
                              ymode=log,ymin=1e2,ymax=3e7,
                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000, 1e10}, 
                              yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$, $10^{10}$},
                              xlabel={Monolithic proof (clauses)}, ylabel={Structural proof + check (clauses)},
            ]

    \input{data-formatted/sub10M-clauses-subthresh}
    \input{data-formatted/sub10M-clauses-supthresh}
    \legend{
      \scriptsize \textsf{Tree ratio $\leq \; 5{.}0$},
      \scriptsize \textsf{Tree ratio $> \; 5{.}0$},
    }
    \addplot[mark=none, color=lightblue] coordinates{(100,100) (2e7,2e7)};
    \addplot[mark=none, color=lightblue] coordinates{(100,2e7) (2e7,2e7)};
    \addplot[mark=none, color=lightblue] coordinates{(2e7,100) (2e7,2e7)};
    \addplot[color=black,dashed] coordinates{(100,2.5e7) (2.5e7,2.5e7)};
    \addplot[color=black,dashed] coordinates{(2.5e7,100) (2.5e7,2.5e7)};

          \end{axis}
\end{tikzpicture}
} % centering
\caption{Structural (Y-axis) versus monolithic (X-axis) proof sizes.  Monolithic generally performed better for lower tree ratios.}
\label{fig:clauses:mono:structured}
\end{figure}

Figure~\ref{fig:clauses:mono:structured} shows the comparative proof
sizes (in clauses) for the two approaches.  As can be seen, the
monolithic approach tends to generate shorter proofs.  Of the 80 cases
for which both methods completed within 1000 seconds, 18 had shorter
proofs with the structural approach and 62 with the monolithic
approach.  In setting our tree ratio threshold to 5.0, we get the shorter proofs for 45 cases and suboptimal proof sizes for 40.

Our summary data is based on the hybrid approach.

\begin{itemize}
\item Of the 123 benchmarks, 110 were completed using the hybrid approach within 10,000 seconds.
\item One additional one completed within 10,000 for structural (and for hybrid with a threshold of $10^5$ clauses).
\item For seven others, was able to do one-sided proof
\item Only one had more than $10^9$ defining clauses.
\item Total of 5 (including 1 overflow case) for which no form of validation succeeded.
  \begin{itemize}
  \item Three of these had more than $10^9$ defining clauses
  \item Smallest had 405 million

Figure~\ref{fig:d4:cpog} compares the running time for the combination
of generating and checking the CPOG file (Y axis), versus the running
time for C4 (X Axis).  If we compare the ratio between the two we find the following:
\begin{itemize}
\item Min = 0.27 (i.e., verification chain was $3.64\times$ faster)
\item Max = 177.0
\item Median = 12.5
\end{itemize}

@@
